---
title: "class19"
author: "Kai Zhao(PID:A17599942)"
format: pdf
---


```{r}

#1 echo=FALSE

cdc <- data.frame( year = c(1922L,1923L,1924L,1925L,1926L,
                                         1927L,1928L,1929L,1930L,1931L,
                                         1932L,1933L,1934L,1935L,1936L,1937L,
                                         1938L,1939L,1940L,1941L,1942L,
                                         1943L,1944L,1945L,1946L,1947L,1948L,
                                         1949L,1950L,1951L,1952L,1953L,1954L,
                                         1955L,1956L,1957L,1958L,1959L,
                                         1960L,1961L,1962L,1963L,1964L,1965L,
                                         1966L,1967L,1968L,1969L,1970L,
                                         1971L,1972L,1973L,1974L,1975L,1976L,
                                         1977L,1978L,1979L,1980L,1981L,
                                         1982L,1983L,1984L,1985L,1986L,1987L,
                                         1988L,1989L,1990L,1991L,1992L,1993L,
                                         1994L,1995L,1996L,1997L,1998L,
                                         1999L,2000L,2001L,2002L,2003L,2004L,
                                         2005L,2006L,2007L,2008L,2009L,
                                         2010L,2011L,2012L,2013L,2014L,2015L,
                                         2016L,2017L,2018L,2019L,2020L,
                                         2021L), cases = c(107473,164191,165418,152003,
                                         202210,181411,161799,197371,166914,
                                         172559,215343,179135,265269,180518,
                                         147237,214652,227319,103188,183866,
                                         222202,191383,191890,109873,133792,
                                         109860,156517,74715,69479,120718,
                                         68687,45030,37129,60886,62786,
                                         31732,28295,32148,40005,14809,11468,
                                         17749,17135,13005,6799,7717,9718,
                                         4810,3285,4249,3036,3287,1759,
                                         2402,1738,1010,2177,2063,1623,1730,
                                         1248,1895,2463,2276,3589,4195,
                                         2823,3450,4157,4570,2719,4083,6586,
                                         4617,5137,7796,6564,7405,7298,
                                         7867,7580,9771,11647,25827,25616,
                                         15632,10454,13278,16858,27550,18719,
                                         48277,28639,32971,20762,17972,
                                         18975,15609,18617,6124,2116)
      )
```

Lets have a wee look at this table

```{r}
head(cdc)
```

>Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(ggplot2)

ggplot(cdc, aes(x = year, y = cases)) +
  geom_point() +  
  geom_line() +  
  labs(title = "CDC Pertussis Case Numbers Over Time",
       x = "Year",
       y = "Number of Cases")

```

>Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
library(ggplot2)

# Your existing plot code
ggplot(cdc, aes(x = year, y = cases)) +
  geom_point() +
  geom_line() +
  labs(title = "CDC Pertussis Case Numbers Over Time",
       x = "Year",
       y = "Number of Cases") +

  geom_vline(xintercept = 1946, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 1996, linetype = "dashed", color = "red") +

  geom_text(aes(x = 1946, y = max(cases), label = "wP vaccine introduced", vjust = -1), color = "blue") +
  geom_text(aes(x = 1996, y = max(cases), label = "Switch to aP vaccine", vjust = -1), color = "red")

```

>Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

###From the lab hands-on The answer is : It is clear from the CDC data that pertussis cases are once again increasing. For example, we can see that in 2012 the CDC reported 48,277 cases of pertussis in the United States. This is the largest number of cases reported since 1955, when 62,786 cases were reported. The pertussis field has several hypotheses for the resurgence of pertussis including (in no particular order): 1) more sensitive PCR-based testing, 2) vaccination hesitancy 3) bacterial evolution (escape from vaccine immunity), 4) waning of immunity in adolescents originally primed as infants with the newer aP vaccine as compared to the older wP vaccine.


### More people get the  diseases.

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```

>Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

```{r}
library(jsonlite)

# Subject table
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector = TRUE)
titer <- read_json("http://cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = TRUE )
```

>Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

>Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

```{r}
library(lubridate)
today()
today() - ymd("2000-01-01")
time_length( today() - ymd("2000-01-01"),  "years")
```
>Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?


```{r}
library(dplyr)
library(lubridate)

subject$age <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)

subject$age_years <- time_length(subject$age,"years")

```

```{r}
ap <- subject %>% filter(infancy_vac == "aP")
round(summary(time_length(ap$age, "years")))
```

```{r}
wp <- subject %>% filter(infancy_vac == "wP")
round(summary(time_length(wp$age, "years")))
```

>Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

>Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
library(ggplot2)
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

```{r}
# Or use wilcox.test() 
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```

>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}

meta <- inner_join(specimen,subject)
dim(meta)
head(meta)
```

>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)

```

>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```
>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```


>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:



```{r}
ggplot(igg)+
  aes(MFI_normalised ,
      antigen) +
  geom_boxplot() +
  xlim(0,75) +
 facet_wrap(vars(visit),nrow = 2)
```

