---
title: "lab12part2extracredits"
author: "Kai Zhao (PID:A17599942)"
format: pdf
---

##Section 4: Population Scale Analysis [HOMEWORK]
One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 expression.



>Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.
How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
summary(expr)
```


```{r}
table(expr$geno)
```
>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Yes
```{r}
library(ggplot2)
```
let make a boxplot
```{r}
ggplot(expr)+aes(geno,exp,fill=geno)+geom_boxplot(notch=TRUE)
```

