---
title: "Class9 helloween"
author: "Kai Zhao (PID: A17599942)"
format: pdf
    
---

Today is Helloween and we will apply lots of the analysis methods ad R graphics approaches to find out all about typical Halloween candy.

```{r}
candy_file <- "candy-data.csv"

candy <- read.csv(candy_file, row.names=1)

head(candy)


```
>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)

```

I can covert the 1 and 0 values to be True and False and use that to extract the typeof candy I want. For example the cholocate candy...
```{r}
candy[as.logical(candy$chocolate),]
```
>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Nerds",]$winpercent
```
>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```


>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

>Q7. What do you think a zero and one represent for the candy$chocolate column?


>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)
ggplot(candy)+
  aes(winpercent)+
  geom_histogram(binwidth=10)
```

>Q9. Is the distribution of winpercent values symmetrical?

```{r}

```
>Q10. Is the center of the distribution above or below 50%?

```{r}

```
>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.inds <- as.logical(candy$chocolate)
choc.win <- candy[choc.inds,]$winpercent
choc.win 
```
>Q12. Is this difference statistically significant?

```{r}
# Do the same for fruity
fruity.inds <- as.logical(candy$fruity)
fruity.win <- candy[fruity.inds,]$winpercent
fruity.win 
```

```{r}
mean(choc.win)
mean(fruity.win)
```
```{r}
t.test(choc.win,fruity.win)
```

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruity.win)
```

> Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(desc(winpercent)) %>% head(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy)+
  aes(winpercent, rownames(candy)) +
  geom_col(fill="blue")
  
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

##Define some useful colors

```{r}
mycols <- rep("blue", nrow(candy))
#mycols[2:5] <- "red"
mycols [as.logical(candy$fruity)] <- "red"
mycols
```

```{r}
#1 fig-height:10
#1 fig-width:5
ggplot(candy)+
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill=mycols)

```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

## Define some useful colars
>Q17. What is the worst ranked chocolate candy?

Sixlet
>Q18. What is the best ranked fruity candy?

Starburst is the best ranked fruity candy. 



library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 15)


>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )

```
> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```


```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```


library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")



```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```


>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().




```{r}



ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
>Q23. Similarly, what two variables are most positively correlated?

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```


```{r}

pca <- prcomp(candy, scale=TRUE)

summary(pca)
pca$rotation[,1]
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", main="PCA Score Plot of PC1 vs PC2")
plot(pca$x[,1:2], col=my_cols, pch=16)
```


```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```



```{r}
library(plotly)
ggplotly(p)
```
>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
